Performance for the masses: Experiments with A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-01-26 GECCO/volunteer-clouds-gecco16.tex
Anonymizing a couple of references, the only ones I found that can discover us. References #95

%Mario: Is NodIO presented here after the cut? 
%TO DO: Add reference
There is, then, space for frameworks that do not make any of these
assumptions, such as {\sf NodIO} presented in \cite{2016arXiv160101607M}. {\sf NodIO} is basically 
%assumptions, such as the one presented in this paper, which will be
% named {\sf NodIO}. {\sf NodIO} is basically
% (Paloma) Is the "first" word really needed? 
% (Mario) I guess is more related to fundamentally or primarily but "basically" sounds better
a two tier client/server 
application, in which clients can run on a browser. 
Therefore, they can be embedded in any web-enabled device, including many over which 
developers have no control.

@@

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. 
%% We need to only reference this section?:
%%------------------- BEGIN
% This paper % (Paloma) No really need for that "also", plus it complicates reading
% establishes the desirable features of
% any {\em global computing} infrastructure, that can also be applied to
% distributed evolutionary computation experiments that rely on
% volunteer computing: scalability, heterogeneity, fault tolerance,
% adaptive parallelism, volunteer's safety, anonymity, hierarchy, ease of use, and
% reasonable performance. But hierarchy, this is, the capability of
% preferentially using resources in the same organization, is a feature
% that could be either disregarded or included within adaptivity, which
% leaves us with eight desirable properties. 
% % (Paloma) Re-write. For example, "But is hierarchy, this is, the
% % capability of preferentially using resources in the same
% % organization, the one that could be either disregarded or included
% % in adaptivity, which leaves us with eight desirable properties." 
% % (Paloma) And I re-wrote what I proposed yesterday.
% % thanks :-) -JJ
% %%------------------- END
% %% EDIT:
% %%------------------- BEGIN
% The fulfillment of these properties could be achieved by using any of the
% languages and associated technologies mentioned above, provided that the user had
% installed the required virtual machine plugin first. 
% %%------------------- END
The fulfillment of the desirable features of any distributed evolutionary 
computation experiments that rely on volunteer computing \cite{2016arXiv160101607M} could be achieved by using any of the
languages and associated technologies, mentioned above, provided that the user had
installed the required virtual machine plugin first.

@@

As far as we know, this paper presents one of the few experiments using % (Paloma) Change to "which" in order to not repeating "that" all the time
computational resources that are as dissimilar as smartphones and
powerful laptops 
or workstation computers in a research center. 
%The methodology used
% (Paloma) I would say "the used methodology"
The algorithms used, as well as the methodology 
for gathering resources and baseline results that we will use for
comparison have been described in
\cite{2016arXiv160101607M}. We will focus next on the experimental
results using the first version of the framework.

@@

Initial results using the same code base and a single computer have
been described in \cite{2016arXiv160101607M}. They prove that a bigger population
contributes to diversity and thus speeds up the solution
\cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing 
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic, although they can
complement them. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which these runs could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the month of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:summary:os}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. There was no special intention to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is indeed negligible.
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of PUTs per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccccc}
\hline
Date & Experiments &\#IPs &\#IPs &  \#IPs & Time (s) &  \#PUTs & $<$ 69s & $<$ 3.46s & Inter-experiment\\
     & \# & Different & Median & Max & Median &  Median & & & correlation \\ 
\hline
4/4 & 57 & 191 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 & 231 & 559 & 5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 97 & 179 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of IPs
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP PUTs per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even in this case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen in 1 out of roughly 5 or roughly 20 cases, depending on what baseline
experiment you choose, it will not happen always or even most of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found. This is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. Thus, the user can decide if to 
continue sharing resources or not. % (Paloma) Feature or disadvantage?
However, it puts a probably            % (Mario) Feature if you only want to share your CPU a little  
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

@@

The results reported above indicated that we should look for a way of
increasing the time that every user devotes. This led us to create an
architecture that uses Web Workers, which is described in
\cite{2016arXiv160101607M}; web workers work in the background even if
the web page is not in focus. A few improvements were also tested,
basically using a random population and restarting the web worker when
the solution was found.

@@

The fact that the solution of this particular problem was found in a
few seconds and restarts were necessary yielded interesting results,
but we needed to work on a harder problem, that is why created a new
client version that optimized Rastrigin function, a multimodal
floating-point problem used as a benchmark in evolutionary
algorithms. The set up for this function, and a comparison of the
baseline speed for JavaScript in an standalone experiment and other
languages is shown in \cite{2016arXiv160101607M}, where the details of
the implementation have also been extensively described.

>>>>

%Mario: Is NodIO presented here after the cut? 
%TO DO: Add reference
There is, then, space for frameworks that do not make any of these
assumptions, such as {\sf NodIO} presented in \cite{2016arXiv160101607Manom}. {\sf NodIO} is basically 
%assumptions, such as the one presented in this paper, which will be
% named {\sf NodIO}. {\sf NodIO} is basically
% (Paloma) Is the "first" word really needed? 
% (Mario) I guess is more related to fundamentally or primarily but "basically" sounds better
a two tier client/server 
application, in which clients can run on a browser. 
Therefore, they can be embedded in any web-enabled device, including many over which 
developers have no control.

@@

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. 
%% We need to only reference this section?:
%%------------------- BEGIN
% This paper % (Paloma) No really need for that "also", plus it complicates reading
% establishes the desirable features of
% any {\em global computing} infrastructure, that can also be applied to
% distributed evolutionary computation experiments that rely on
% volunteer computing: scalability, heterogeneity, fault tolerance,
% adaptive parallelism, volunteer's safety, anonymity, hierarchy, ease of use, and
% reasonable performance. But hierarchy, this is, the capability of
% preferentially using resources in the same organization, is a feature
% that could be either disregarded or included within adaptivity, which
% leaves us with eight desirable properties. 
% % (Paloma) Re-write. For example, "But is hierarchy, this is, the
% % capability of preferentially using resources in the same
% % organization, the one that could be either disregarded or included
% % in adaptivity, which leaves us with eight desirable properties." 
% % (Paloma) And I re-wrote what I proposed yesterday.
% % thanks :-) -JJ
% %%------------------- END
% %% EDIT:
% %%------------------- BEGIN
% The fulfillment of these properties could be achieved by using any of the
% languages and associated technologies mentioned above, provided that the user had
% installed the required virtual machine plugin first. 
% %%------------------- END
The fulfillment of the desirable features of any distributed evolutionary 
computation experiments that rely on volunteer computing \cite{2016arXiv160101607Manom} could be achieved by using any of the
languages and associated technologies, mentioned above, provided that the user had
installed the required virtual machine plugin first.

@@

As far as we know, this paper presents one of the few experiments using % (Paloma) Change to "which" in order to not repeating "that" all the time
computational resources that are as dissimilar as smartphones and
powerful laptops 
or workstation computers in a research center. 
%The methodology used
% (Paloma) I would say "the used methodology"
The algorithms used, as well as the methodology 
for gathering resources and baseline results that we will use for
comparison have been described in
\cite{2016arXiv160101607Manom}. We will focus next on the experimental
results using the first version of the framework.

@@

Initial results using the same code base and a single computer have
been described in \cite{2016arXiv160101607Manom}. They prove that a bigger population
contributes to diversity and thus speeds up the solution
\cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing 
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic, although they can
complement them. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which these runs could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15anom}. For the
purpose of this paper, we repeated the announcement several times
through the month of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:summary:os}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. There was no special intention to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is indeed negligible.
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of PUTs per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccccc}
\hline
Date & Experiments &\#IPs &\#IPs &  \#IPs & Time (s) &  \#PUTs & $<$ 69s & $<$ 3.46s & Inter-experiment\\
     & \# & Different & Median & Max & Median &  Median & & & correlation \\ 
\hline
4/4 & 57 & 191 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 & 231 & 559 & 5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 97 & 179 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of IPs
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP PUTs per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15anom} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even in this case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen in 1 out of roughly 5 or roughly 20 cases, depending on what baseline
experiment you choose, it will not happen always or even most of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found. This is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. Thus, the user can decide if to 
continue sharing resources or not. % (Paloma) Feature or disadvantage?
However, it puts a probably            % (Mario) Feature if you only want to share your CPU a little  
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

@@

The results reported above indicated that we should look for a way of
increasing the time that every user devotes. This led us to create an
architecture that uses Web Workers, which is described in
\cite{2016arXiv160101607Manom}; web workers work in the background even if
the web page is not in focus. A few improvements were also tested,
basically using a random population and restarting the web worker when
the solution was found.

@@

The fact that the solution of this particular problem was found in a
few seconds and restarts were necessary yielded interesting results,
but we needed to work on a harder problem, that is why created a new
client version that optimized Rastrigin function, a multimodal
floating-point problem used as a benchmark in evolutionary
algorithms. The set up for this function, and a comparison of the
baseline speed for JavaScript in an standalone experiment and other
languages is shown in \cite{2016arXiv160101607Manom}, where the details of
the implementation have also been extensively described.
